Assessing Bridge Conditions in New York State: The Impact of Age, Municipality, and Structural Deficiencies

### 2024-11-29 paper/paper.qmd
Add revision of Data section and upload draft of Model Section.

## Outcome Variable
### Condition 
The **Condition** variable reflects the overall state of a bridge, encompassing both its structural integrity and functional adequacy as determined during inspections. It is measured on a continuous scale from 1 to 7, with higher scores indicating better condition. Bridges with condition scores above 5 are considered to be in good standing, while scores below 5 indicate that a bridge may be classified as either Structurally Deficient (SD) or Functionally Obsolete (FO).

@@

# Summary statistics for Condition
condition_summary <- bridge_data %>%
  summarise(
    Min = min(Condition, na.rm = TRUE),
    Max = max(Condition, na.rm = TRUE),
    Mean = mean(Condition, na.rm = TRUE),
    Median = median(Condition, na.rm = TRUE),
    "Standard Deviation" = sd(Condition, na.rm = TRUE),
    Count = n()
  )

# Create a table
kable(condition_summary, format = "latex", booktabs = TRUE)
```
@tbl-condition-summary presents for the condition ratings of bridges in New York State, based on a dataset of 15,855 observations. The mean condition rating of 5.38 indicates that, on average, bridges are in fair to good condition. The median rating of 5.30 suggests that half of the bridges have condition ratings below or equal to 5.30. The close alignment of the mean and median reflects a relatively symmetrical distribution. While the minimum condition rating is 1.93, and the maximum is 7, the standard deviation of 0.83 suggests that most bridges cluster around the mean, with only a few outliers at the extreme ends of the scale. These statistics indicate that the majority of bridges are maintained at a stable and adequate level, with relatively few in very poor or excellent condition.

@@

# Create a histogram for the Condition variable
ggplot(bridge_data, aes(x = Condition)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(1, 7)) + # Ensures x-axis shows 1 to 7
  labs(
    x = "Condition",
    y = "Frequency"
  ) +
  theme_minimal()
```

@@

## Predictor variables
### AgeAtInspection
The **AgeAtInspection** variable is a continuous measure that represents the age of a bridge in years. It is calculated as the time elapsed from the year the bridge was originally built or entirely replaced to its most recent inspection date.

@@

# Visualization: Histogram with mode markers and density curve
ggplot(bridge_data, aes(x = AgeAtInspection)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black", alpha = 0.7) + # Histogram
  geom_vline(xintercept = seq(25, 150, by = 25), color = "red", linetype = "dashed", size = 0.8) + # Vertical lines for modes
  geom_density(aes(y = ..count..), color = "blue", size = 1) + # Density curve scaled to histogram
  labs(
    x = "Age at Inspection (Years)",
    y = "Frequency"
  ) +
  theme_minimal()

```
@fig-age-histogram shows distribution of bridge ages at the time of inspection, overlaid with red dashed lines and a blue density curve. The red dashed lines mark intervals of 25 years and align with clear peaks in the histogram, suggesting periodic patterns in bridge construction or major renovations. Prominent clusters are observed around 25, 50, and 75 years of age, indicating that many bridges were likely built or reconstructed during these timeframes. The blue density curve provides a smooth visualization of the distribution, confirming the periodic modes while highlighting the overall trends. Beyond 100 years, the frequency drops significantly, with very few bridges older than this still in operation, reflecting the typical lifecycle of infrastructure. The concentration of bridges between 20 and 80 years old suggests that a significant portion of bridges are mid-life and may soon require maintenance or replacement.@fig-age-histogram highlights both the historical patterns of bridge construction and the current age distribution of bridges in the dataset.
```{r}
#| label: fig-ageatinspection-condition
#| fig-cap: "Relationship Between AgeAtInspection and Condition. Each blue dot represents a bridge, showing its age at the time of inspection and the corresponding condition rating obtained from that inspection."
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Visualization: Relationship between AgeAtInspection and Condition
ggplot(bridge_data, aes(x = AgeAtInspection, y = Condition)) +
  geom_point(alpha = 0.3, color = "blue") +
  labs(
    x = "Age at Inspection (Years)",
    y = "Condition"
  ) +
  theme_minimal()
```
@fig-ageatinspection-condition illustrates the relationship between bridge age at inspection and their condition ratings. Most points are concentrated in the upper-left quadrant, showing that younger bridges tend to have higher condition ratings (close to 7). As bridge age increases, condition ratings generally decline, reflecting the expected deterioration over time. However, beyond approximately 80 to 100 years, the relationship becomes less defined, with condition ratings displaying a wide range. This variability may be influenced by differences in maintenance practices, varying levels of transportation load, or disparities in construction quality among older bridges. Overall, the pattern indicates that bridge age is a key factor in determining condition, particularly for younger and midlife bridges.

### Located_Municipality
The **Located_Municipality** variable categorizes bridges based on the type of municipality where they are located. The three categories are City, Town, and Village.

```{r}
#| label: tbl-located-municipality-proportion
#| tbl-cap: "Proportion of Bridges by Municipality. The first column shows the type of municipality (Located_Municipality), and the second column represents the proportion of bridges in each category (Percentage Account)."
#| echo: false
#| tbl-pos: "H"

@@

# Create a formatted table
kable(located_municipality_proportion, format = "latex", booktabs = TRUE)
```
@tbl-located-municipality-proportion shows the distribution of bridges across different municipality types. Towns account for the majority of the bridges, representing 82.11% of the dataset, while cities and villages make up much smaller proportions, at 9.15% and 8.75%, respectively. This indicates that most bridges are located in towns, likely due to their larger geographic coverage.
```{r}
#| label: tbl-located-municipality-condition
#| tbl-cap: "Average Bridge Condition by Municipality. The first column shows the type of municipality (Located_Municipality), and the second column provides the average condition rating (Average Condition) for bridges in each category."
#| echo: false
#| tbl-pos: "H"

# Average condition by municipality
located_municipality_condition <- bridge_data %>%
  group_by(Located_Municipality) %>%
  summarise(
    "Average Condition" = round(mean(Condition, na.rm = TRUE), 2)
  )

@@

### Owner_Group
The **Owner_Group** variable categorizes bridges based on their ownership. This variable is designed to estimate potential differences in bridge conditions that may arise from varying ownership responsibilities. The categories include Municipalities, NYSDOT, and Other. The Other category combines bridges owned by private entities, authorities, commissions, and similar groups, as each accounts for a small portion of the data, making it practical to group them together
```{r}
#| label: tbl-owner-group-proportion
#| tbl-cap: "Proportion of Bridges by Ownership. The first column shows the type of ownership group (Owner_Group), and the second column represents the proportion of bridges in each category (Percentage Account)."
#| echo: false
#| tbl-pos: "H"

@@

@@

```{r}
#| label: tbl-owner-group-condition
#| tbl-cap: "Average Bridge Condition by Ownership. The first column shows the type of ownership group (Owner_Group), and the second column provides the average condition rating (Average Condition) for bridges in each category."
#| echo: false
#| tbl-pos: "H"

# Average condition by ownership
owner_group_condition <- bridge_data %>%
  group_by(Owner_Group) %>%
  summarise(
    "Average Condition" = round(mean(Condition, na.rm = TRUE), 2)
  )

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

>>>>

## Outcome Variable
### Condition 
The **Condition** variable reflects the overall state of a bridge, encompassing both its structural integrity and functional adequacy as determined during inspections. It is measured on a continuous scale from 1 to 7, with higher scores indicating better condition. Bridges with condition scores above 5 are considered to be in good standing, while scores below 5 indicate that a bridge is classified as either Structurally Deficient (SD) or Functionally Obsolete (FO).

@@

# Summary statistics for Condition
condition_summary <- bridge_data %>%
  summarise(
    Min = round(min(Condition, na.rm = TRUE), 2),
    Max = round(max(Condition, na.rm = TRUE), 2),
    Mean = round(mean(Condition, na.rm = TRUE), 2),
    Median = round(median(Condition, na.rm = TRUE), 2),
    "Standard Deviation" = round(sd(Condition, na.rm = TRUE), 2),
    Count = n()
  )

# Create a table
kable(condition_summary, format = "latex", booktabs = TRUE)
```
@tbl-condition-summary presents for the condition ratings of bridges in New York State, based on a clenaed dataset of 15,855 observations. The mean condition rating of 5.38 indicates that, on average, bridges are in fair to good condition. The median rating of 5.30 suggests that half of the bridges have condition ratings below or equal to 5.30. The close alignment of the mean and median reflects a relatively symmetrical distribution. While the minimum condition rating is 1.93, and the maximum is 7, the standard deviation of 0.83 suggests that most bridges cluster around the mean, with only a few outliers at the extreme ends of the scale. These statistics indicate that the majority of bridges are maintained at a stable and adequate level, with relatively few in very poor or excellent condition.

@@

# Create a histogram for the Condition variable
ggplot(bridge_data, aes(x = Condition)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(1, 7)) + # Ensures x-axis shows 1 to 7
  labs(
    x = "Bridge Condition Rating",
    y = "Frequency"
  ) +
  theme_minimal()
```

@@

## Predictor variables
### Age of Bridge at Inspection
The **AgeAtInspection** variable is a continuous measure that represents the age of a bridge in years. It is calculated as the time elapsed from the year the bridge was originally built or entirely replaced to its most recent inspection date.

@@

# Visualization: Histogram with mode markers and density curve
ggplot(bridge_data, aes(x = AgeAtInspection)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black", alpha = 0.7) + # Histogram
  geom_vline(xintercept = seq(25, 150, by = 25), color = "red", linetype = "dashed", size = 0.8) + # Vertical lines for modes
  geom_density(aes(y = ..count..), color = "blue", size = 1) + # Density curve scaled to histogram
  labs(
    x = "Bridge Ages at Inspection (Years)",
    y = "Frequency"
  ) +
  theme_minimal()

```
@fig-age-histogram shows distribution of bridge ages at the time of inspection, overlaid with red dashed lines and a blue density curve. The red dashed lines mark intervals of 25 years and align with clear peaks in the histogram, suggesting periodic patterns in bridge construction or major renovations. Prominent clusters are observed around 25, 50, and 75 years of age, indicating that many bridges were likely built or reconstructed during these timeframes. The blue density curve provides a smooth visualization of the distribution, confirming the periodic modes while highlighting the overall trends. Beyond 100 years, the frequency drops significantly, with very few bridges older than this still in operation, reflecting the typical lifecycle of infrastructure. The concentration of bridges between 20 and 80 years old suggests that a significant portion of bridges are mid-life and may soon require maintenance or replacement.It highlights both the historical patterns of bridge construction and the current age distribution of bridges in the dataset.
```{r}
#| label: fig-ageatinspection-condition
#| fig-cap: "Relationship Between Bridge Age at Inspection and Bridge Condition Rating. Each blue dot represents a bridge, showing its age at the time of inspection and the corresponding condition rating obtained from that inspection."
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

# Visualization: Relationship between AgeAtInspection and Condition
ggplot(bridge_data, aes(x = AgeAtInspection, y = Condition)) +
  geom_point(alpha = 0.3, color = "blue") +
  labs(
    x = "Bridge Age at Inspection (Years)",
    y = "Bridge Condition Rating"
  ) +
  theme_minimal()
```
@fig-ageatinspection-condition illustrates the relationship between bridge age at inspection and their condition ratings. Most points are concentrated in the upper-left quadrant, showing that younger bridges tend to have higher condition ratings (close to 7). As bridge age increases, condition ratings generally decline, reflecting the expected deterioration over time. However, beyond approximately 80 to 100 years, the relationship becomes less defined, with condition ratings displaying a wide range. This variability may be influenced by differences in maintenance practices, varying levels of transportation load, or disparities in construction quality among older bridges. Overall, the pattern indicates that bridge age is a key factor in determining condition, particularly for younger and midlife bridges.

### Bridge Located Municipality
The **Located_Municipality** variable categorizes bridges based on the type of municipality where they are located. The three categories are City, Town, and Village.

```{r}
#| label: tbl-located-municipality-proportion
#| tbl-cap: "Proportion of Bridges by Municipality. The first column lists the type of municipality where the bridges are located, including City, Town, and Village, while the second column shows the proportion of bridges in each category."
#| echo: false
#| tbl-pos: "H"

@@

# Rename the first column of the summarize table
colnames(located_municipality_proportion)[
  colnames(located_municipality_proportion) == "Located_Municipality"
] <- "Bridge Located Municipality"

# Create a formatted table
kable(located_municipality_proportion, format = "latex", booktabs = TRUE)
```
@tbl-located-municipality-proportion shows the distribution of bridges across different municipality types. Towns account for the majority of the bridges, representing 82.11% of the dataset, while cities and villages make up much smaller proportions, at 9.15% and 8.75%, respectively. This indicates that most bridges are located in towns, likely due to their larger geographic coverage.
```{r}
#| label: tbl-located-municipality-condition
#| tbl-cap: "Average Bridge Condition Rating by Municipality. The first column lists the type of municipality where the bridges are located, including City, Town, and Village, and the second column provides the average condition rating for bridges in each category."
#| echo: false
#| tbl-pos: "H"

# Average condition by municipality
located_municipality_condition <- bridge_data %>%
  group_by(Located_Municipality) %>%
  summarise(
    "Average Condition Rating" = round(mean(Condition, na.rm = TRUE), 2)
  )

# Rename the first column of the summarize table
colnames(located_municipality_condition)[
  colnames(located_municipality_condition) == "Located_Municipality"
] <- "Bridge Located Municipality"

@@

### Ownership of Bridge
The **Owner_Group** variable categorizes bridges based on their ownership. This variable is designed to estimate potential differences in bridge conditions that may arise from varying ownership responsibilities. The categories include Municipalities, NYSDOT, and Other. The Other category combines bridges owned by private entities, authorities, commissions, and similar groups, as each accounts for a small portion of the data, making it practical to group them together
```{r}
#| label: tbl-owner-group-proportion
#| tbl-cap: "Proportion of Bridges by Ownership. The first column lists the type of bridge ownership, including NYSDOT, Municipality, and Other (private entities, authorities, commissions), while the second column shows the proportion of bridges in each category."
#| echo: false
#| tbl-pos: "H"

@@

# Rename the first column of the summarize table
colnames(owner_group_proportion)[
  colnames(owner_group_proportion) == "Owner_Group"
] <- "Ownership of Bridge"

@@

```{r}
#| label: tbl-owner-group-condition
#| tbl-cap: "Average Bridge Condition Rating by Ownership. The first column lists the type of bridge ownership, including NYSDOT, Municipality, and Other (private entities, authorities, commissions), and the second column provides the average condition rating for bridges in each category."
#| echo: false
#| tbl-pos: "H"

# Average condition by ownership
owner_group_condition <- bridge_data %>%
  group_by(Owner_Group) %>%
  summarise(
    "Average Condition Rating" = round(mean(Condition, na.rm = TRUE), 2)
  )

# Rename the first column of the summarize table
colnames(owner_group_condition)[
  colnames(owner_group_condition) == "Owner_Group"
] <- "Ownership of Bridge"

@@

The goal of our modeling strategy is to investigate the relationship between bridge condition ratings and key predictors, including **AgeAtInspection**, **Located_Municipality**, and **Owner_Group**. The model aims to identify significant patterns and evaluate how these factors influence bridge conditions.

Here, we present a linear regression model, explaining its setup, assumptions, and justification. Details of model diagnostics and validation can be found in [Appendix -@sec-model-details].

## Model Setup

Define $y_i$ as the condition rating of bridge $i$, measured on a scale from 1 to 7, where higher values represent better condition. The predictors are defined as follows:
- $x_{1i}$: Age of bridge $i$ at the time of inspection (**AgeAtInspection**), measured in years.
- $x_{2i}$: Type of municipality where the bridge is located (**Located_Municipality**), categorized as **City**, **Town**, or **Village**.
- $x_{3i}$: Type of ownership of bridge $i$ (**Owner_Group**), categorized as **Municipality**, **NYSDOT**, or **Other**.

The linear regression model is expressed as:
```{=latex}
\[
\begin{aligned}
y_i &= \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i, \text{City}} + \beta_3 x_{2i, \text{Village}} \\
&\quad + \beta_4 x_{3i, \text{NYSDOT}} + \beta_5 x_{3i, \text{Other}} + \epsilon_i, \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2),
\end{aligned}
\]

The model assumes that the relationship between predictors and the response variable is linear, and that the residuals are normally distributed with constant variance.

### Software Implementation

We implemented the model in R using the `lm()` function, which fits a linear regression model using ordinary least squares. Model validation and diagnostics, including residual analysis and goodness-of-fit metrics, were conducted to assess model performance.

## Model Justification

The inclusion of **AgeAtInspection** as a continuous predictor reflects its importance in capturing the natural deterioration of bridges over time. **Located_Municipality** and **Owner_Group** were included as categorical predictors to account for variations in maintenance practices, funding, and infrastructure quality that differ across municipalities and ownership types. These variables align directly with those defined in the data section, ensuring consistency between the data description and modeling approach.

We chose a linear regression model for its simplicity, interpretability, and appropriateness for the continuous response variable (**Condition**). While other models, such as generalized additive models or Bayesian regressions, could capture non-linearities or incorporate priors, the current dataset supports a linear approach without overcomplicating the analysis.

### Assumptions and Limitations

The model assumes:
- A linear relationship between predictors and the response variable.
- Homoscedasticity (constant variance of residuals).
- Normality of residuals.
- Independence of observations.

Potential limitations include the exclusion of interaction terms and other factors (e.g., traffic volume, environmental conditions) that could influence bridge condition. Furthermore, the linear regression model may not capture complex non-linear relationships or spatial dependencies inherent in the data.

### Validation and Diagnostics

To validate the model, we performed:
- **Residual Analysis**: Checked for patterns in residuals to confirm assumptions of linearity and homoscedasticity.
- **Goodness-of-Fit Metrics**: Calculated $R^2$ and adjusted $R^2$ to evaluate how well the model explains variability in the data.
- **Sensitivity Analysis**: Assessed the impact of excluding extreme outliers on the estimated coefficients.
- **Alternative Models**: Considered models with interaction terms and different transformations of predictors but found no substantial improvement in model performance.

Details of model diagnostics, including plots of residuals and additional goodness-of-fit metrics, are provided in [Appendix -@sec-model-details].

@@
